---
title: "Housing-related Census Data in the Region of Waterloo"
author: "Craig A. Sloss"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(sf)
library(leaflet)
library(RColorBrewer)
census = readRDS("./data/selected_census_records.rds")
percentage_renter_dataset = census %>%
  filter(CHARACTERISTIC_NAME == "Renter") %>%
  select(GEO_NAME, pct_renter = C10_RATE_TOTAL)
percentage_high_housing_cost_dataset = census %>%
  filter(CHARACTERISTIC_NAME == "Spending 30% or more of income on shelter costs") %>%
  select(GEO_NAME, pct_high_housing_cost = C10_RATE_TOTAL)
da_lookup = readRDS("./data/da_lookup_table.rds")
waterloo_region_map = read_sf(dsn = "./data/da_shapefile", layer = "lda_000b21a_e") %>%
  st_transform("+proj=longlat +datum=WGS84") %>%
  left_join(da_lookup) %>%
  filter(CDNAME == "Waterloo") %>%
  left_join(percentage_renter_dataset %>% rename(DAUID = GEO_NAME)) %>%
  left_join(percentage_high_housing_cost_dataset %>% rename(DAUID = GEO_NAME)) %>%
  mutate(combined_renter_and_high_cost = pct_renter * pct_high_housing_cost / 100)
```

#### Percentage of Renters in Waterloo Region

```{r}
pal <- colorNumeric(
  palette = "Purples",
  domain = waterloo_region_map$pct_renter
)
leaflet(waterloo_region_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = waterloo_region_map,
              weight = 0,
              color = ~pal(pct_renter),
              fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~pct_renter, title = "") 
```

#### Percentage of Residents Paying More than 30% or more of Income on Shelter

Note that this includes both owners and tenants.

```{r}
pal <- colorNumeric(
  palette = "Purples",
  domain = waterloo_region_map$pct_high_housing_cost
)
leaflet(waterloo_region_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = waterloo_region_map,
              weight = 0,
              color = ~pal(pct_high_housing_cost),
              fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~pct_high_housing_cost, title = "") 
```

#### Combining Rental Rates and High Housing Cost Rates

This map combines the percentage of renters and the percentage of residents with high housing costs into a single metric that identifies areas where both percentages are high. It should not be interpreted directly as a percentage.

```{r}
pal <- colorNumeric(
  palette = "Purples",
  domain = waterloo_region_map$combined_renter_and_high_cost
)
leaflet(waterloo_region_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = waterloo_region_map,
              weight = 0,
              color = ~pal(combined_renter_and_high_cost),
              fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~combined_renter_and_high_cost, title = "") 
```

#### Data sources 

* Statistics Canada, 2021 Census of Population (https://www12.statcan.gc.ca/census-recensement/2021/dp-pd/prof/details/download-telecharger.cfm?Lang=E)

* Statistics Canada, 2021 Census Boundaries (https://www12.statcan.gc.ca/census-recensement/2021/geo/sip-pis/boundary-limites/index2021-eng.cfm?Year=21)

