---
title: "Housing-related Census Data in the Region of Waterloo"
author: "Craig A. Sloss"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
census = readRDS("./data/selected_census_records.rds")
da_lookup = readRDS("./data/da_lookup_table.rds")
percentage_renter_dataset = census %>%
  filter(CHARACTERISTIC_NAME == "Renter") %>%
  select(GEO_NAME, pct_renter = C10_RATE_TOTAL)
rental_weight_dataset = census %>%
  filter(CHARACTERISTIC_NAME == "Total - Private households by tenure - 25% sample data") %>%
  select(GEO_NAME, rental_weight = C1_COUNT_TOTAL)
population_dataset = census %>%
  filter(CHARACTERISTIC_NAME == "Population, 2021") %>%
  select(GEO_NAME, population = C1_COUNT_TOTAL)
combined_dataset = percentage_renter_dataset %>%
  left_join(population_dataset) %>%
  left_join(rental_weight_dataset) %>%
  left_join(da_lookup %>% rename(GEO_NAME = DAUID))
```

#### Percentage of Renters by City in Region of Waterloo

Note: census percentages of renters are an estimate based on 25% of the population. Therefore the numbers below will be an approximation of the true number.

```{r}
combined_dataset %>%
  filter(CDNAME == "Waterloo") %>%
  group_by(CSDNAME) %>%
  summarise(total_population = sum(population, na.rm = TRUE),
            pct_renter_estimate = round(sum(rental_weight * pct_renter, na.rm = TRUE) / sum(rental_weight, na.rm = TRUE), 1),
            renter_count_estimate = round(pct_renter_estimate * total_population / 100, 0)) %>%
  rename(Municipality = CSDNAME,
         `Total Population` = total_population,
         `Renters -- Estimated Percent` = pct_renter_estimate,
         `Renters -- Estimated Count` = renter_count_estimate) %>%
  knitr::kable()
```

#### Percentage of Renters -- Region of Waterloo vs Other Census Divisions in Ontario

Note: census percentages of renters are an estimate based on 25% of the population. Therefore the numbers below will be an approximation of the true number.

```{r}
combined_dataset %>%
  filter(!is.na(CDNAME)) %>%
  group_by(CDNAME) %>%
  summarise(total_population = sum(population, na.rm = TRUE),
            pct_renter_estimate = round(sum(rental_weight * pct_renter, na.rm = TRUE) / sum(rental_weight, na.rm = TRUE), 1),
            renter_count_estimate = round(pct_renter_estimate * total_population / 100, 0)) %>%
  arrange(-renter_count_estimate) %>%
  rename(`Census Division` = CDNAME,
         `Total Population` = total_population,
         `Renters -- Estimated Percent` = pct_renter_estimate,
         `Renters -- Estimated Count` = renter_count_estimate) %>%
  knitr::kable()
```