---
title: "Workshop -- Exploring Census Mapping Data"
author: "Craig A. Sloss"
date: "2023-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Run this code to load the packages used in this notebook:
```{r}
library(tidyverse)
library(sf)
library(leaflet)
```

Download the Census Cartographic Dissemination Area Shapefile from here: https://www12.statcan.gc.ca/census-recensement/2021/geo/sip-pis/boundary-limites/index2021-eng.cfm?Year=21

Unzip the file and place the folder in the project directory

Load the file

# Visualize the file

Create a map with just a base layer using the addProviderTiles function

Type help(addProviderTiles) to explore some of the options available

Import the Census DA file

Plot the outline of the DA boundaries on the map

# Identify Waterloo Region

To find out which DAs correspond to each city, download the geography relationship file here: https://www12.statcan.gc.ca/census-recensement/2021/geo/sip-pis/dguid-idugd/index2021-eng.cfm?year=21

Place the file in the project folder

Import the file into R

Work through understanding this sample code:

```{r}
da_list = data.frame(DAUID = shapefile$DAUID, DGUID = shapefile$DGUID)
csd_list = data.frame(DGUID = census_subdivision$DGUID, CSDNAME = census_subdivision$CSDNAME)
cd_list = data.frame(DGUID = census_division$DGUID, CDNAME = census_division$CDNAME)
relationship_file_unique = relationship_file %>% 
  distinct(DADGUID_ADIDUGD, CSDDGUID_SDRIDUGD, CDDGUID_DRIDUGD) 
lookup_table = da_list %>%
  left_join(relationship_file_unique, by = c("DGUID" = "DADGUID_ADIDUGD")) %>%
  left_join(csd_list, by = c("CSDDGUID_SDRIDUGD" = "DGUID")) %>%
  left_join(cd_list, by = c("CDDGUID_DRIDUGD" = "DGUID"))
waterloo_region_lookup = lookup_table %>%
  filter(CDNAME == "Waterloo") %>%
  select(DAUID, CSDNAME)
saveRDS(waterloo_region_lookup, "./waterloo_region_DA_lookup.rds")
```

Join the lookup file to the shape file. 

Filter so that only one city appears.

Replot the map.

Save the shapefile for later use.